---
title: "V3"
author: "Teddy"
date: "2025-03-30"
output: html_document
---

```{r setup, include=FALSE}
# install.packages("imputeTS")
# install.packages("randomForest")
# install.packages("imputeTS")
# install.packages("gbm")
# install.packages("scales")
# install.packages("forecast")
# install.packages("imputeTS")
# install.packages("glmnet")
# install.packages("rugarch")
# install.packages("mgcv")
# install.packages("psych")
# install.packages("corrplot")
# install.packages("reshape2")
library(reshape2)
library(corrplot)
library(mgcv)
library(tidyverse)
library(lubridate)
library(randomForest)
library(gbm)
library(ggplot2)
library(scales)
library(forecast)
library(imputeTS)
library(glmnet)
library(rugarch)
library(psych)
library(zoo)
library(Metrics)
```

### Data preparation 

```{r}
# Load
cocoa_prices <- read.csv("Daily Prices_ICCO.csv", stringsAsFactors = FALSE)
cocoa_prices$Price <- as.numeric(gsub(",", "", cocoa_prices$ICCO.daily.price..US..tonne.))
cocoa_prices$Date <- as.Date(cocoa_prices$Date, format='%d/%m/%Y')

# Clean
cocoa_prices <- cocoa_prices %>%
  select(Date, Price) %>%
  arrange(Date) %>%
  filter(!is.na(Price) & Price > 0)

# Compute log and differenced log
cocoa_prices <- cocoa_prices %>%
  mutate(
    log_price = log(Price),
    diff_log_price = c(NA, diff(log_price))
  )
```

```{r}
# Load
ghana_weather <- read.csv("Ghana_data.csv", stringsAsFactors = FALSE)
ghana_weather$DATE <- as.Date(ghana_weather$DATE)

ghana_weather <- ghana_weather %>%
  group_by(DATE) %>%
  summarise(
    PRCP = mean(PRCP, na.rm = TRUE),
    TAVG = mean(TAVG, na.rm = TRUE),
    TMAX = mean(TMAX, na.rm = TRUE),
    TMIN = mean(TMIN, na.rm = TRUE),
    .groups = "drop"
  )

# Clean
date_seq <- tibble(DATE = seq(min(ghana_weather$DATE), max(ghana_weather$DATE), by = "day"))

# Join and fill missing days with NA
ghana_weather <- date_seq %>%
  left_join(ghana_weather, by = "DATE")

# Fill NA values using Kalman smoothing for daily series
ghana_weather <- ghana_weather %>%
  mutate(
    PRCP = na_kalman(PRCP, model = "StructTS", smooth = TRUE),
    TAVG = na_kalman(TAVG, model = "StructTS", smooth = TRUE),
    TMAX = na_kalman(TMAX, model = "StructTS", smooth = TRUE),
    TMIN = na_kalman(TMIN, model = "StructTS", smooth = TRUE)
  )

# Aggregate to monthly level and plot
cocoa_prices_monthly <- cocoa_prices %>%
  mutate(YearMonth = floor_date(Date, "month")) %>%
  group_by(YearMonth) %>%
  summarise(
    AvgPrice = mean(Price, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    log_price = log(AvgPrice),
    diff_log_price = c(NA, diff(log_price))
  )
```

### Tables & Descriptive plots
```{r}
# Histogram of cocoa price
ggplot(cocoa_prices, aes(x = Price)) +
  geom_histogram(binwidth = 100, fill = "steelblue", color = "white") +
  labs(title = "Histogram of Cocoa Price", x = "Price (USD/tonne)", y = "Count") +
  theme_minimal()

# Histogram of log-transformed price
ggplot(cocoa_prices, aes(x = log_price)) +
  geom_histogram(binwidth = 0.1, fill = "darkorange", color = "white") +
  labs(title = "Histogram of Log Price", x = "Log(Price)", y = "Count") +
  theme_minimal()

# Histogram of precipitation
ggplot(ghana_weather, aes(x = PRCP)) +
  geom_histogram(binwidth = 0.1, fill = "skyblue", color = "white") +
  labs(title = "Histogram of Precipitation (PRCP)", x = "PRCP", y = "Count") +
  theme_minimal()

# Histogram of average temperature
ggplot(ghana_weather, aes(x = TAVG)) +
  geom_histogram(binwidth = 1, fill = "tomato", color = "white") +
  labs(title = "Histogram of Avg Temperature (TAVG)", x = "TAVG (°F)", y = "Count") +
  theme_minimal()

# Histogram of max temperature
ggplot(ghana_weather, aes(x = TMAX)) +
  geom_histogram(binwidth = 1, fill = "seagreen", color = "white") +
  labs(title = "Histogram of Max Temperature (TMAX)", x = "TMAX (°F)", y = "Count") +
  theme_minimal()

# Histogram of min temperature
ggplot(ghana_weather, aes(x = TMIN)) +
  geom_histogram(binwidth = 1, fill = "purple", color = "white") +
  labs(title = "Histogram of Min Temperature (TMIN)", x = "TMIN (°F)", y = "Count") +
  theme_minimal()
```


### Merge
```{r}
cocoa_data <- left_join(cocoa_prices, ghana_weather, by = c("Date" = "DATE")) %>% mutate(log_price = log(Price),
         diff_log_price = c(NA, diff(log_price))) %>% na.omit()
```

### Plotting
```{r}
diff_log_ts <- diff(log(cocoa_data$Price))

ggplot(cocoa_data, aes(x = Date, y = Price)) +
  geom_line(color = "blue", size = 0.8) +
  scale_y_continuous(labels = dollar_format(prefix = "$")) +
  labs(title = "Daily Cocoa Prices", x = NULL, y = "Price (USD)") +
  theme_minimal(base_size = 13)

ggplot(cocoa_data, aes(x = Date, y = log_price)) +
  geom_line(color = "darkgreen", size = 0.8) +
  labs(title = "Log Transformed Cocoa Prices", x = NULL, y = "Log(Price)") +
  theme_minimal(base_size = 13)

plot(cocoa_prices_monthly$YearMonth, cocoa_prices_monthly$diff_log_price,
     type = "l", col = "blue",
     main = "Differenced Monthly Log Cocoa Prices",
     ylab = "Diff Log(Price)", xlab = "Date")

# ACF plot
acf(diff_log_ts, main = "ACF of Differenced Log Cocoa Prices", col = "blue", lwd = 2)

# PACF plot
pacf(diff_log_ts, main = "PACF of Differenced Log Cocoa Prices", col = "darkgreen", lwd = 2)

# Month-wise price trend
cocoa_prices_monthly$Month <- month(cocoa_prices_monthly$YearMonth, label = TRUE)
ggplot(cocoa_prices_monthly, aes(x = Month, y = AvgPrice)) +
  geom_boxplot(fill = "skyblue") +
  labs(title = "Seasonal Trend of Cocoa Prices (Monthly)", x = "Month", y = "Avg Price") +
  theme_minimal()

# Volatility plot
cocoa_prices_monthly$abs_diff <- abs(cocoa_prices_monthly$diff_log_price)
ggplot(cocoa_prices_monthly, aes(x = YearMonth, y = abs_diff)) +
  geom_line(color = "orange", size = 0.7) +
  labs(title = "Volatility of Monthly Cocoa Prices", y = "|Δlog(Price)|", x = "Month") +
  theme_minimal()

# STL Decomposition
ts_log_price <- ts(cocoa_prices_monthly$log_price, frequency = 12, start = c(year(min(cocoa_prices_monthly$YearMonth)), month(min(cocoa_prices_monthly$YearMonth))))
decomp <- stl(ts_log_price, s.window = "periodic")
autoplot(decomp) +
  labs(title = "STL Decomposition of Monthly Log Cocoa Prices") +
  theme_minimal()

# Seasonal plot
ggseasonplot(ts_log_price, year.labels = TRUE, year.labels.left = TRUE) +
  labs(title = "Seasonal Plot of Monthly Log Cocoa Prices", y = "Log(Price)", x = "Month") +
  theme_minimal()

# Subseries plot
ggsubseriesplot(ts_log_price) +
  labs(title = "Subseries Plot of Monthly Log Cocoa Prices", y = "Log(Price)", x = "Month") +
  theme_minimal()
```

### Some addtional plots
```{r}
# Correlation Plot
num_vars <- cocoa_data %>% select(where(is.numeric))
corrplot(cor(num_vars, use = "complete.obs"), method = "color", type = "upper", tl.cex = 0.8)

# Missing Value Heatmap
ggplot(cocoa_data, aes(x = Date)) +
  geom_point(aes(y = Price), alpha = 0.3) +
  labs(title = "Price Data Over Time (Missing Check)", y = "Price") +
  theme_minimal()

# Smoothed Trend
ggplot(cocoa_data, aes(x = Date, y = Price)) +
  geom_line(alpha = 0.4) +
  geom_smooth(method = "loess", color = "red") +
  labs(title = "Smoothed Trend of Cocoa Prices", x = "Date", y = "Price") +
  theme_minimal()
```


### Split Data into Training and Testing Sets
```{r}
set.seed(6657)
train_size <- floor(0.8 * nrow(cocoa_data))
train_data <- cocoa_data[1:train_size, ]
test_data <- cocoa_data[(train_size + 1):nrow(cocoa_data), ]
```

### Choosing predictors
```{r}
cor(cocoa_data[, c("log_price", "diff_log_price", "TAVG", "PRCP", "TMAX", "TMIN")], use = "complete.obs")
```
### Volatility Check Plot
```{r}
# Compute log returns
log_returns <- diff(log(cocoa_data$Price))
log_returns <- na.omit(log_returns)
dates <- cocoa_data$Date[2:(length(log_returns) + 1)]

# Plot log returns
plot(dates, log_returns, type = "l", col = "darkblue",
     main = "Log Returns of Cocoa Prices",
     xlab = "Date", ylab = "Log Return")

# Plot squared log returns (volatility clustering check)
squared_returns <- log_returns^2
plot(dates, squared_returns, type = "l", col = "firebrick",
     main = "Squared Log Returns (Check for Volatility Clustering)",
     xlab = "Date", ylab = expression(r[t]^2))

# ACF of squared returns
acf(squared_returns, main = "ACF of Squared Log Returns")

# Ljung-Box test on squared returns
ljung_result <- Box.test(squared_returns, lag = 10, type = "Ljung-Box")
print(ljung_result)
```

### ARIMA model
```{r}
ts_data <- ts(cocoa_prices_monthly$AvgPrice, start = c(1994, 10), frequency = 12)

arima_model <- Arima(ts_data, order = c(0,1,1), include.mean = TRUE)

# Extract fitted values
fitted_values <- fitted(arima_model)
plot(ts_data, main = "Observed vs. Fitted Values (ARIMA(0,1,1))", col = "blue", type = "l", ylab = "Diff  Price", xlab = "Time")

# Add the fitted values in red
lines(fitted_values, col = "red", lwd = 2)

# Add legend
legend("topright", legend = c("Observed", "Fitted"), col = c("blue", "red"), lty = 1, lwd = 2)

arima_model <- Arima(ts_data, order = c(0,1,1), include.mean = TRUE)

# Forecast next 4 months
forecast_values <- forecast(arima_model, h = 4)

# Get time index
time_index <- time(ts_data)

# Filter data from 2024 onward
start_year <- 2024
ts_subset <- window(ts_data, start = c(start_year, 1))
fitted_subset <- window(fitted(arima_model), start = c(start_year, 1))

# Create future time points for forecast (monthly assumed)
future_time <- time(ts_data)[length(ts_data)] + seq(1, 4)/12

# Plot observed data (since 2020)
plot(ts_subset, 
     main = "Observed, Fitted & Forecasted Values (ARIMA(0,1,1)) Since 2024",
     col = "blue", lwd = 2, ylab = "Price", xlab = "Time",
     xlim = c(start_year, future_time[4]), ylim = range(c(ts_subset, forecast_values$mean, forecast_values$lower[,2], forecast_values$upper[,2])))

# Add fitted values
lines(fitted_subset, col = "red", lwd = 2)

# Add forecasted values
lines(future_time, as.numeric(forecast_values$mean), col = "green", lwd = 2, lty = 2)

# Add 95% confidence interval
lines(future_time, forecast_values$lower[,2], col = "black", lty = 5)
lines(future_time, forecast_values$upper[,2], col = "black", lty = 5)

# Add legend
legend("topleft",
       legend = c("Observed (since 2024)", "Fitted", "Forecast", "95% CI"),
       col = c("blue", "red", "green","black"),
       lwd = 2, lty = c(1, 1, 2, 3))

```


### GAM model
```{r}
# Create lag features
cocoa_lagged <- cocoa_data |>
  mutate(
    lag_1 = lag(log_price, 1),
    lag_2 = lag(log_price, 2),
    date_int = as.numeric(Date)  # Convert Date to numeric for smoothing
  ) |>
  filter(
    !is.na(log_price),
    !is.na(lag_1),
    !is.na(lag_2),
    !is.na(TAVG),
    !is.na(TMAX)
  )

# Split the data
set.seed(8848)
split_index <- floor(0.8 * nrow(cocoa_lagged))
data_train <- cocoa_lagged[1:split_index, ]
data_test <- cocoa_lagged[(split_index + 1):nrow(cocoa_lagged), ]

# Fit GAM model
my_gam <- gam(log_price ~ s(date_int) + lag_1 + lag_2 + TAVG + TMAX,
              data = data_train, method = "ML")

h <- 3000 
last_row <- tail(data_train, 1)
future_forecast <- tibble(
  Date = seq(last_row$Date + 1, by = "days", length.out = h),
  lag_1 = NA_real_,
  lag_2 = NA_real_,
  TAVG = rep(mean(data_train$TAVG, na.rm = TRUE), h),
  TMAX = rep(mean(data_train$TMAX, na.rm = TRUE), h),
  date_int = as.numeric(seq(last_row$Date + 1, by = "days", length.out = h))
)

# Initialize lags
lag_1 <- last_row$log_price
lag_2 <- last_row$lag_1
log_forecasts <- numeric(h)

# walk-forward
for (i in 1:h) {
  new_data <- tibble(
    date_int = future_forecast$date_int[i],
    lag_1 = lag_1,
    lag_2 = lag_2,
    TAVG = future_forecast$TAVG[i],
    TMAX = future_forecast$TMAX[i]
  )
  pred_log <- predict(my_gam, newdata = new_data)
  log_forecasts[i] <- pred_log
  lag_2 <- lag_1
  lag_1 <- pred_log
}

future_forecast$Price <- exp(log_forecasts)
future_forecast$Model <- "GAM"

ggplot() +
  geom_line(data = cocoa_data, aes(x = Date, y = Price), color = "black") +
  geom_line(data = future_forecast, aes(x = Date, y = Price, color = Model), linewidth = 1.2) +
  scale_color_manual(values = c("GAM" = "orange")) +
  labs(
    title = "GAM Forecast vs Actual Prices (Walk-Forward Forecast)",
    x = "Date",
    y = "Price",
    color = "Model"
  ) +
  theme_minimal()
```

### GARCH and ARCH
```{r}
# Compute log returns
returns <- cocoa_data %>%
  mutate(log_return = log(Price) - lag(log(Price))) %>%
  drop_na()

# Fit GARCH(1,1) model
garch <- ugarchspec(
  variance.model = list(model = "sGARCH", garchOrder = c(1, 1)),
  mean.model = list(armaOrder = c(0, 0), include.mean = TRUE),
  distribution.model = "norm"
)

fit_garch <- ugarchfit(spec = garch, data = returns$log_return)

g_1 <- tibble(
  Date = returns$Date,
  g_1 = as.numeric(sigma(fit_garch))
)
r_1 <- rollapply(abs(returns$log_return), 
                 width = 5, FUN = sd, 
                 align = "right", fill = NA)

r_df <- tibble(
  Date = returns$Date,
  r_1 = r_1
) %>% drop_na()

vol_df <- g_1 %>%
  inner_join(r_df, by = "Date")

# Plot
ggplot(vol_df, aes(x = Date)) +
  geom_line(aes(y = g_1, color = "GARCH Estimated"), linewidth = 0.8) +
  geom_line(aes(y = r_1, color = "Realized (Rolling SD)"), linewidth = 0.8) +
  scale_color_manual(values = c("GARCH Estimated" = "red", "Realized (Rolling SD)" = "yellow")) +
  labs(
    title = "GARCH Estimated vs Actual Volatility",
    x = "Date",
    y = "Volatility",
    color = "Type"
  ) +
  theme_minimal()

### Evaluations

# Standardized residuals
resid_std <- residuals(fit_garch, standardize = TRUE)

# Coefficient table
fit_garch@fit$matcoef

# Stationarity check
alpha1 <- fit_garch@fit$coef["alpha1"]
beta1 <- fit_garch@fit$coef["beta1"]
sum_ab <- alpha1 + beta1
cat("alpha + beta =", round(sum_ab, 4), "\n")

if (sum_ab < 1) {
  cat("Stationary: yes\n")
} else {
  cat("Stationary: no\n")
}

# Ljung-Box test
Box.test(resid_std, lag = 10, type = "Ljung-Box")
Box.test(resid_std^2, lag = 10, type = "Ljung-Box")
```


### Walk-Forward Lag Selection for RF and GBM Forecasting
```{r}
# This section tunes the number of lag features (1 to 5) for forecasting cocoa prices.
# It uses repeated walk-forward validation on the last 150 observations of the dataset.
# For each lag setting, it fits both a Random Forest and a GBM model to predict
# the difference in log prices (diff_log_price), then transforms predictions back
# to raw price scale to calculate RMSE. The average RMSE across repetitions is used.

# Creating lag 1-5
create_lags <- function(data, lags = 1:7) {
  set.seed(10086)
  lag_vars <- lapply(lags, function(l) dplyr::lag(data$log_price, l))
  names(lag_vars) <- paste0("lag_", lags)
  lag_df <- as.data.frame(lag_vars)
  dplyr::bind_cols(data, lag_df)
}

# RMSE holder
rmse_results <- data.frame(Lags = integer(), RF_RMSE = numeric(), GBM_RMSE = numeric())

# last 150 rows for tuning
cocoa_data_short <- cocoa_data[(nrow(cocoa_data) - 150):nrow(cocoa_data), ]

N_rep <- 5
for (lag_n in 1:5) {
  rmse_rf_vec <- c()
  rmse_gbm_vec <- c()

  for (rep in 1:N_rep) {
    set.seed(100 + rep)
    cocoa_data_lagged <- create_lags(cocoa_data_short, lags = 1:lag_n)
    cocoa_data_lagged <- cocoa_data_lagged %>%
      mutate(diff_log_price = c(NA, diff(log_price))) %>%
      drop_na()

    model_data <- cocoa_data_lagged
    start_index <- nrow(model_data) - 30
    pred_rf <- c()
    pred_gbm <- c()
    actual <- model_data$Price[(start_index+1):nrow(model_data)]

    for (i in start_index:(nrow(model_data) - 1)) {
      train_data <- model_data[1:i, ]
      test_data <- model_data[i+1, ]

      predictors <- c(paste0("lag_", 1:lag_n), "TAVG", "TMAX")
      formula <- as.formula(paste("diff_log_price ~", paste(predictors, collapse = "+")))

      set.seed(42 + rep)
      rf_model <- randomForest(formula, data = train_data)

      set.seed(42 + rep)
      gbm_model <- gbm(formula, data = train_data, distribution = "gaussian",
                       n.trees = 100, interaction.depth = 3, shrinkage = 0.05,
                       bag.fraction = 1, verbose = FALSE)

      rf_pred <- predict(rf_model, newdata = test_data)
      gbm_pred <- predict(gbm_model, newdata = test_data, n.trees = 100)

      last_log_price <- log(model_data$Price[i])
      rf_back <- exp(last_log_price + rf_pred)
      gbm_back <- exp(last_log_price + gbm_pred)

      pred_rf <- c(pred_rf, rf_back)
      pred_gbm <- c(pred_gbm, gbm_back)
    }
    rmse_rf_vec <- c(rmse_rf_vec, sqrt(mean((pred_rf - actual)^2)))
    rmse_gbm_vec <- c(rmse_gbm_vec, sqrt(mean((pred_gbm - actual)^2)))
  }
  rmse_results <- rbind(rmse_results,
                        data.frame(Lags = lag_n,
                                   RF_RMSE = mean(rmse_rf_vec),
                                   GBM_RMSE = mean(rmse_gbm_vec)))
}

# Find best lag based on average RMSE
best_rf_lag <- rmse_results$Lags[which.min(rmse_results$RF_RMSE)]
best_gbm_lag <- rmse_results$Lags[which.min(rmse_results$GBM_RMSE)]
```

```{r}
# Plotting results
results_long <- melt(rmse_results, id.vars = "Lags", variable.name = "Model", value.name = "RMSE")
ggplot(results_long, aes(x = Lags, y = RMSE, color = Model)) +
  geom_line() +
  geom_point() +
  labs(title = "Average RMSE by Number of Lags (Repeated Walk-Forward)",
       x = "Number of Lags", y = "Average RMSE")
```


### Expanding Window Cross-Validation Function (RF_model1 and GBM_model1)
```{r}
# This function performs expanding window cross-validation for time series forecasting.
# It takes in lagged features and fits either a Random Forest or GBM model.
# The model is retrained in each iteration as more data becomes available.
# Forecasts are made for one step ahead repeatedly and predictions are returned
# in a tibble with date, actual price, and predicted price.

expanding_window_cv <- function(data, model_type = "rf", lag = 5, step = 7) {
  step <- as.numeric(step)
  if (nrow(data) < 30) {
    stop("No enough data") # check enough
  }

  preds <- c()
  actuals <- c()
  dates <- c()

  init_size <- floor(0.8 * nrow(data))
  formula <- as.formula(paste("log_price ~", paste0("lag_", 1:lag, collapse = " + "), " + TAVG + TMAX"))

  for (i in seq(init_size, nrow(data) - 1, by = step)) {
    train <- data[1:i, ]
    test <- data[i + 1, , drop = FALSE]

    if (any(is.na(test)) || any(is.na(train))) next

    set.seed(7727)

    model <- if (model_type == "rf") {
      randomForest(formula, data = train, ntree = 50)
    } else {
      gbm(formula, data = train, distribution = "gaussian",
          n.trees = 30, shrinkage = 0.05, interaction.depth = 3,
          bag.fraction = 1, verbose = FALSE)
    }
    pred_log <- if (model_type == "rf") {
      predict(model, test)
    } else {
      predict(model, test, n.trees = 30)
    }
    preds <- c(preds, exp(pred_log))
    actuals <- c(actuals, test$Price)
    dates <- c(dates, test$Date)
  }

  return(tibble(Date = dates, Actual = actuals, Predicted = preds))
}

final_lagged_data <- create_lags(cocoa_data, lags = 1:max(best_rf_lag, best_gbm_lag)) %>%
  mutate(log_price = log(Price)) %>%
  drop_na()
```

### Final RF and GBM Forecasting Using Optimal Lag Variables
### *** May take 10 to 15 mins *** :(
```{r}
# Run RF and GBM forecasts using best lag values(10 mins)
rf_df <- expanding_window_cv(final_lagged_data, model_type = "rf", lag = best_rf_lag, step = 7)
gbm_df <- expanding_window_cv(final_lagged_data, model_type = "gbm", lag = best_gbm_lag, step = 7)
```

### Evaluation
```{r}
# Plot actual vs predicted
plot_df <- bind_rows(
  rf_df %>% mutate(Model = "RF"),
  gbm_df %>% mutate(Model = "GBM")
)
ggplot() +
  geom_line(data = plot_df, aes(x = Date, y = Actual), color = "black") +
  geom_line(data = plot_df, aes(x = Date, y = Predicted, color = Model), linewidth = 0.6) +
  labs(title = "Expanding Window Forecast (RF & GBM)", x = "Date", y = "Price") +
  theme_minimal()

#Residuals
plot_df <- bind_rows(
  rf_df %>% mutate(Model = "RF", Residual = Actual - Predicted),
  gbm_df %>% mutate(Model = "GBM", Residual = Actual - Predicted)
)

ggplot(plot_df, aes(x = Date, y = Residual, color = Model)) +
  geom_line() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Prediction Residuals Over Time", y = "Residual (Actual - Predicted)", x = "Date") +
  theme_minimal()

ggplot(plot_df, aes(x = Actual, y = Predicted, color = Model)) +
  geom_point(alpha = 0.4) +
  geom_abline(slope = 1, intercept = 0, color = "black", linetype = "dashed") +
  labs(title = "Actual vs Predicted Prices", x = "Actual Price", y = "Predicted Price") +
  theme_minimal()

# RF performance
rf_rmse <- rmse(rf_df$Actual, rf_df$Predicted)
rf_mae <- mae(rf_df$Actual, rf_df$Predicted)
rf_mape <- mape(rf_df$Actual, rf_df$Predicted)

# GBM performance
gbm_rmse <- rmse(gbm_df$Actual, gbm_df$Predicted)
gbm_mae <- mae(gbm_df$Actual, gbm_df$Predicted)
gbm_mape <- mape(gbm_df$Actual, gbm_df$Predicted)

results <- tibble(
  Model = c("RF", "GBM"),
  RMSE = c(rf_rmse, gbm_rmse),
  MAE = c(rf_mae, gbm_mae),
  MAPE = c(rf_mape, gbm_mape)
)

# Residual Histogram
ggplot(plot_df, aes(x = Residual, fill = Model)) +
  geom_histogram(alpha = 0.6, position = "identity", bins = 30) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(title = "Residual Distribution of RF and GBM Forecasts",
       x = "Residual (Actual - Predicted)",
       y = "Count") +
  theme_minimal()
print(results)

acf(rf_df$Actual - rf_df$Predicted, main = "ACF of RF Forecast Residuals")
```

```{r}
# RMSE plot reliability
rf_results <- rf_df %>%
  mutate(Error = Predicted - Actual,
         SE = Error^2,
         Step = row_number(),
         RMSE = sqrt(SE))

ggplot(rf_results, aes(x = Step, y = RMSE)) +
  geom_line(color = "steelblue") +
  geom_point(size = 1) +
  labs(title = "Walk-forward RMSE of Random Forest Forecasts",
       x = "Forecast Step",
       y = "RMSE") +
  theme_minimal()
```

### Forecast Next 2000 Days Using Random Forest (walk-forward)
```{r}
predict_next_days_rf <- function(data, model_lag = 5, days_ahead = 90, ntree = 50) {
  if (nrow(data) < 30 || !"log_price" %in% names(data)) {
    stop("Data must contain log_price and be at least 30 rows.")
  }

  last <- tail(data, 1)

  rf_formula <- as.formula(
    paste("log_price ~", paste0("lag_", 1:model_lag, collapse = " + "), " + TAVG + TMAX")
  )

  rf_model <- randomForest(
    formula = rf_formula,
    data = data,
    ntree = ntree
  )

  future_preds <- tibble()
  current_input <- last

  for (i in 1:days_ahead) {
    pred_log <- predict(rf_model, newdata = current_input)
    pred_price <- exp(pred_log)

    new_row <- current_input %>%
      mutate(
        Date = current_input$Date + days(1),
        Price = pred_price,
        log_price = pred_log
      )

    for (j in model_lag:2) {
      new_row[[paste0("lag_", j)]] <- current_input[[paste0("lag_", j - 1)]]
    }
    new_row[["lag_1"]] <- pred_log

    new_row$TAVG <- current_input$TAVG + rnorm(1, mean = 0, sd = 0.5)
    new_row$TMAX <- current_input$TMAX + rnorm(1, mean = 0, sd = 0.5)

    future_preds <- bind_rows(future_preds, new_row)
    current_input <- new_row
  }

  return(future_preds)
}

```

```{r}
set.seed(1314)
future_rf_2000 <- predict_next_days_rf(
  data = final_lagged_data,
  model_lag = best_rf_lag,
  days_ahead = 2000
)
```

```{r}
plot_df <- bind_rows(
  final_lagged_data %>% select(Date, Price) %>% mutate(Type = "Observed"),
  future_rf_2000  %>% select(Date, Price) %>% mutate(Type = "Forecasted")
)

ggplot(plot_df, aes(x = Date, y = Price, color = Type)) +
  geom_line(size = 0.9) +
  geom_vline(xintercept = max(final_lagged_data$Date), linetype = "dashed", color = "black") +
  scale_color_manual(values = c("Observed" = "#1f77b4", "Forecasted" = "#e41a1c")) +
  labs(title = "2000-Day Forecast (RF Model)", x = "Date", y = "Price (USD)") +
  theme_minimal()
```



